Greenhouse gas emissions model

### 2021-08-25 210617_GHG_model.Rmd
attempt at cleaning up the code, added PLS, plot all distributions (all the things in the return), use the uncertainty tools on the total C02 and milk yield.. Still a lot to do.. @eikeluedeling PTAL (probably enough just to open the html)

#Automatically write R package citation entries to a .bib file
knitr::write_bib(c(.packages(), 
                   'decisionSupport'), 'references/packages.bib')

```

@@

We're looking at the equation put together by Joshua Wafula to check if this can easily be used by decisionSupport [@R-decisionSupport], and to possibly optimize the model.

The formulas come from Chapter 10 of the *2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@dong_ipcc_2006] (Volume 4 - Agriculture, Forestry and Other Land Use). The document can be found on [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_10_Ch10_Livestock.pdf). The chapter is part of the book *Good Practice Guidance and Uncertainty Management in National Greenhouse Gas Inventories* [@ipcc_good_2021] , also on the [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/gp/english/). Many variables are based on the work done in *Variation in the carbon footprint of milk production on smallholder dairy farms in central Kenya* [@wilkes_variation_2020] and in *Methods and guidance to support MRV of livestock emissions* [@wilkes_methods_2019].

**Here are our changes and additions to the original model:**

```{r load_data}

library(decisionSupport)
###CW Note ####
# #### remove all the 'C:/Users/jwafula/Dropbox/UNIQUE_CCAFS/analysis/data/' folder locations throughout - 
#### these are not useful with RMD and RStudio / Git integrations
###CW Note ####
### # This project will be stored in relative file paths (i.e. same or lower folders)
input_table <- "data/livestock_ghg_input_table.csv"
```

@@

```{r make_variables}
### Reasonable ranges still missing for many variables
make_variables <- function(est,n=1){ x <- random(rho=est, n=n)
for(i in colnames(x)) assign(i, as.numeric(x[1,i]),envir=.GlobalEnv)}

@@

# CH4 emission factor from manure management (Equation 10.23)
mm_ch4<-(volsol*365)*
        (Bo*0.67*
           ((mms_pasture*MCFpasture)+
            (mms_spread*MCFspread)+
            (mms_drylot*MCFdrylot)+
            (mms_solidstore*MCFsolidstore)+
            (mms_composted*MCFcomposted)+
            (mms_liquid_slurry*MCFliquid)+
            (mms_biogas*MCFbiogas)+
            (mms_burn*MCFburn)+
              # the original code was lacking the 'sold' fraction - we added this for completeness
            (mms_sold*MCFsold)))

@@

# comment: this is probably wrong, since leached nitrogen doesn't emit (much) nitrous oxide
#mm_leach_n2o<-n_excretion*((mms_pasture*0.3)+(mms_spread*0.3)+(mms_drylot*0.3)+(mms_solidstore*0.3)+
#              (mms_composted*0.3)+(mms_liquid*0.3)+(mms_biogas*0)+(mms_burn*0))*0.0075*(44/28)

@@

# Emissions from feed production and transport. 
# comment: the numbers in the table seem spuriously precise.
feed_CO2<-feed_prod_CO2*feed_trans_CO2

@@

#CW Note ####
# Not clear what this part does (through to the ggplot)

# Relationship between greenhouse gas intensity and milk yields. 
# Fat adjusts
#ann_adj <-kg_FPCM_year*365 
#df <- data.frame(
#  GHGi = on_farm/ann_adj, 
#  FPCM = ann_adj)

# ### ##### CW Note
# Removing this plot
# plot

#library(ggplot2)
#ggplot(df, aes(x = FPCM, y = GHGi)) +geom_point() + ylim(0,50)

return(list(enteric_CH4=enteric_CH4,
            #CW NOTE ####
            # ####working backwards from 'enteric_CO2' I see that 
            # #### this is the product of something twice multiplied by 365
            # ####### 
            milk_yield=sum(total_milk),
            enteric_CH4_CO2eq=sum(enteric_CH4_CO2eq),
            mm_CH4_CO2eq=sum(mm_CH4_CO2eq),
            mm_N2O_CO2eq=sum(mm_N2O_CO2eq),
            feed_CO2=feed_CO2,
            on_farm=sum(on_farm)))
}

```

@@

```{r monte_carlo_simulation}

GHG_simulation <- mcSimulation(
  estimate = estimate_read_csv("data/livestock_ghg_input_table.csv"),
  model_function = ghg_emissions,
  numberOfModelRuns = 1e3, #1000 for now
  functionSyntax = "plainNames"
)

```

```{r plot_distributions}

plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "enteric_CH4_CO2eq",
                   method = "boxplot_density",
                   old_names = "enteric_CH4_CO2eq",
                   new_names = "Methane Emissions from Enteric Fermentation")

```

# uncertainty

Here we apply a collection of functions for use in assessing uncertainty from the `uncertainty` package. Install using `devtools::install_github("CWWhitney/uncertainty")`.

@@

```{r uncertainty_package}
uncertainty::varscatter(in_var = GHG_simulation$y$milk_yield,
                        out_var = GHG_simulation$y$enteric_CH4_CO2eq)
```

Use `varkernel()` for another type of graphical representation of the same relationship. `varkernel()` performs a two-dimensional kernel density estimation for the expected value of the outcome variable `out_var` and the influence variable of interest `in_var` using the `kde2d()` function in the `MASS()` package (Venables and Ripley 2002). The function produces a matrix of the estimated density (z) of the expected value of the outcome variable `out_var` (y) and the influence variable of interest `in_var` (x). It looks slightly different than the scatter plot estimates above because the density function restricts the shape of the kernel to a bivariate normal kernel.

Check the effect of `in_var` `milk_yield` on the `out_var` `enteric_CH4_CO2eq`.

```{r name_in_out_vars}
in_var  <- GHG_simulation$y$milk_yield
out_var <- GHG_simulation$y$enteric_CH4_CO2eq
```

@@

```{r varviolin}
uncertainty::varviolin(in_var = in_var, out_var = out_var)
```

@@

`varkernelslicerange()` shows the influencing variable `in_var` value intervals, optimized interquartile ranges shown in `varviolin()` can be used to select a range to slice from the density kernel `varkernel()` as was done for a single variable value in `varkernelslice()`. Here we set the maximum expected value for `in_var` to 4 and the minimum expected value to 1.

>>>>

#Automatically write R package citation entries to a .bib file
knitr::write_bib(c(.packages(), 
                   'DiagrammeR',
                   'decisionSupport',
                   'MASS'), 'references/packages.bib')
```

@@

The formulas come from Chapter 10 of the *2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@dong_ipcc_2006] (Volume 4 - Agriculture, Forestry and Other Land Use). The document can be found on [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_10_Ch10_Livestock.pdf). The chapter is part of the book *Good Practice Guidance and Uncertainty Management in National Greenhouse Gas Inventories* [@ipcc_good_2021], also on the [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/gp/english/). Many variables are based on the work done in *Variation in the carbon footprint of milk production on smallholder dairy farms in central Kenya* [@wilkes_variation_2020] and in *Methods and guidance to support MRV of livestock emissions* [@wilkes_methods_2019]. We run the model with the decisionSupport package [@R-decisionSupport].

We load the library and the data. The data contains ranges of possible values for the simulation. Each variable is represented with a confidence interval range and a specification of the distribution shape.

```{r load_data}
library(DiagrammeR)
library(decisionSupport)
library(tidyverse)
### # This project will be stored in relative file paths (i.e. same or lower folders)
input_table <- "data/livestock_ghg_input_table.csv"
```

The manure management is based on a decision tree.

## Decision tree

Decision tree describing the options for manure management. Each step in the decision tree (each node) represents a share of manure that is managed according to the name of the node. Green arrows indicate 'Yes' and red arrows 'No'. The descriptions for the uses can be found in the IPCC report on manure management chapter 10 [@dong_ipcc_2006].

```{r graph-mermaid-manure, echo=FALSE, fig.width=10, fig.height=1}
mermaid("
        graph LR
        C(Central collection)-->B(Burned); linkStyle 0 stroke: red, stroke-width:1px
                                B-->P(Pasture); linkStyle 1 stroke: red, stroke-width:1px
        C-->D(Daily spread); linkStyle 2 stroke: green, stroke-width:1px
            D-->S(Sold); linkStyle 3 stroke: red, stroke-width:1px
                S-->L(Liquid storage); linkStyle 4 stroke: red, stroke-width:1px
                    L-->Co(Composted); linkStyle 5 stroke: red, stroke-width:1px
                        Co-->Dl(Dry lot); linkStyle 6 stroke: red, stroke-width:1px
                             Dl-->St(Solid storage); linkStyle 7 stroke: red, stroke-width:1px
                    L-->Bi(Biogas); linkStyle 8 stroke: green, stroke-width:1px
                        Bi-->Ls(Liquid slurry); linkStyle 9 stroke: red, stroke-width:1px
        ")
```

@@

```{r make_variables}

make_variables <- function(est,n=1){ x <- random(rho=est, n=n)
for(i in colnames(x)) assign(i, as.numeric(x[1,i]),envir=.GlobalEnv)}

@@

# CH4 emission factor from manure management (Equation 10.23)
mm_ch4<-(volsol*365)*
        (Bo*0.67*
           ((mms_pasture*MCFpasture)+
            (mms_spread*MCFspread)+
            (mms_drylot*MCFdrylot)+
            (mms_solidstore*MCFsolidstore)+
            (mms_composted*MCFcomposted)+
            (mms_liquid_slurry*MCFliquid)+
            (mms_biogas*MCFbiogas)+
            (mms_burn*MCFburn)+
              # CW EL Comment ####
              # the original code was lacking the 'sold' fraction - we added this for completeness
            (mms_sold*MCFsold)))

@@

# CW EL Comment ####
# # comment: this is probably wrong, since leached nitrogen doesn't emit (much) nitrous oxide
#mm_leach_n2o<-n_excretion*((mms_pasture*0.3)+(mms_spread*0.3)+(mms_drylot*0.3)+(mms_solidstore*0.3)+
#              (mms_composted*0.3)+(mms_liquid*0.3)+(mms_biogas*0)+(mms_burn*0))*0.0075*(44/28)

@@

# Emissions from feed production and transport. 
# CW EL Comment ####
  # comment: the numbers in the table seem spuriously precise.
  ####

feed_CO2<-feed_prod_CO2*feed_trans_CO2

@@

return(list(enteric_CH4=enteric_CH4,
            milk_yield=sum(total_milk),
            enteric_CH4_CO2eq=sum(enteric_CH4_CO2eq),
            mm_CH4_CO2eq=sum(mm_CH4_CO2eq),
            mm_N2O_CO2eq=sum(mm_N2O_CO2eq),
            feed_CO2=feed_CO2,
            on_farm=sum(on_farm)))
}
```

@@

```{r monte_carlo_simulation}
GHG_simulation <- mcSimulation(
  estimate = estimate_read_csv("data/livestock_ghg_input_table.csv"),
  model_function = ghg_emissions,
  numberOfModelRuns = 1e2, #100 for now
  functionSyntax = "plainNames"
)
```

Here we plot the distributions of the various outputs from the model.

```{r plot_distributions, fig.width=10, fig.height=10}
# enteric_CH4=enteric_CH4,
enteric_CH4_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "enteric_CH4",
                   method = "boxplot_density",
                   old_names = "enteric_CH4",
                   new_names = "Total Methane Emissions from Livestock by Enteric Fermentation") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

# milk_yield=sum(total_milk),
milk_yield_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "milk_yield",
                   method = "boxplot_density",
                   old_names = "milk_yield",
                   new_names = "Total Milk Yield") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

# enteric_CH4_CO2eq=sum(enteric_CH4_CO2eq),
enteric_CH4_CO2eq_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "enteric_CH4_CO2eq",
                   method = "boxplot_density",
                   old_names = "enteric_CH4_CO2eq",
                   new_names = "Carbon Dioxide Equivalent Enteric Methane Emissions from Livestock") + 
  theme(axis.title.x=element_blank())

# mm_CH4_CO2eq=sum(mm_CH4_CO2eq),
mm_CH4_CO2eq_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "mm_CH4_CO2eq",
                   method = "boxplot_density",
                   old_names = "mm_CH4_CO2eq",
                   new_names = "Carbon Dioxide Equivalent Methane Emissions from Livestock Manure") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

# mm_N2O_CO2eq=sum(mm_N2O_CO2eq),
mm_N2O_CO2eq_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "mm_N2O_CO2eq",
                   method = "boxplot_density",
                   old_names = "mm_N2O_CO2eq",
                   new_names = "Carbon Dioxide Equivalent Nitrous Oxide Emissions from Livestock Manure") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

# feed_CO2=feed_CO2,
feed_CO2_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "feed_CO2",
                   method = "boxplot_density",
                   old_names = "feed_CO2",
                   new_names = "Carbon Dioxide Emissions from Livestock Feed Production and Transport") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

# on_farm=sum(on_farm)
on_farm_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation,
                   vars = "on_farm",
                   method = "boxplot_density",
                   old_names = "on_farm",
                   new_names = "Total Carbon Dioxide Equivalent Emissions from Livestock") + 
  theme(axis.title.y=element_blank())

library(patchwork)
# remove the axis marks from the two lefthand-most plots and the x-axis title from the left and right plots

# plot with patchwork
layout <- "
           AAABBB
           CCCDDD
           EEEFFF
           GGGGGG
          "

enteric_CH4_dist_plot + milk_yield_dist_plot + 
  enteric_CH4_CO2eq_dist_plot + mm_CH4_CO2eq_dist_plot + 
  mm_N2O_CO2eq_dist_plot + feed_CO2_dist_plot + 
  on_farm_dist_plot + 
plot_layout(guides = "collect", design = layout) 
```

# Model sensitivity

#### Perform a Partial Least Squares Regression (PLSR) of Monte Carlo simulation results

We apply a post-hoc analysis to the `mcSimulation()` outputs with `plsr.mcSimulation()` to determine the Variable Importance in the Projection (VIP) score and coefficients of a Projection to Latent Structures (PLS) regression model. This function uses the outputs of the `mcSimulation()` selecting all the input variables from the decision analysis function in the parameter `object` and then runs a PLS regression with an outcome variable defined in the parameter `resultName`. We use the code `names(GHG_simulation$y)[7]` to select the outcome variable `r names(GHG_simulation$y)[7]`, which is an element of the list `y` in our `GHG_simulation` outputs (this must be a character element).

```{r}
pls_result <- plsr.mcSimulation(object = GHG_simulation,
                  resultName = names(GHG_simulation$y)[7], ncomp = 1)
```

We run the `plot_pls()` on the results from `plsr.mcSimulation()` with a number of standard settings. The length of the bars is equal to VIP with a vertical line at '1' on the x-axis indicating a standard cut-off for VIP used for variable selection [@Whitney2017]. The overall plot only shows those variables with a VIP > 0.8, which is the common threshold for variable selection [@Lanzanova2019, @Luedeling2016]. The colors of the bars represent the positive or negative coefficient of the given input variable with the output variable.

Here we import the input table again to replace the labels for the variables on the y-axis. The input table can include a 'label' and 'variable' column. The standard labels (from the 'variable' column) are usually computer readable and not very nice for a plot. The `plot_pls()` function uses the text in the 'label' column as replacement for the default text in the 'variable' column.

```{r}
input_table <- read.csv("data/livestock_ghg_input_table.csv")

plot_pls(pls_result, input_table = input_table, threshold = 0)
```

# Uncertainty

In the following sections we apply a collection of functions for use in assessing uncertainty from the `uncertainty` package. Install using `devtools::install_github("CWWhitney/uncertainty")`.

@@

```{r uncertainty_package}
uncertainty::varscatter(in_var = GHG_simulation$y$milk_yield,
                        out_var = GHG_simulation$y$on_farm)
```

Use `varkernel()` for another type of graphical representation of the same relationship. `varkernel()` performs a two-dimensional kernel density estimation for the expected value of the outcome variable `out_var` and the influence variable of interest `in_var` using the `kde2d()` function in the `MASS()` package [@R-MASS]. The function produces a matrix of the estimated density (z) of the expected value of the outcome variable `out_var` (y) and the influence variable of interest `in_var` (x). It looks slightly different than the scatter plot estimates above because the density function restricts the shape of the kernel to a bivariate normal kernel.

Check the effect of `in_var` `milk_yield` on the `out_var` `on_farm`.

```{r name_in_out_vars}
in_var  <- GHG_simulation$y$milk_yield
out_var <- GHG_simulation$y$on_farm
```

@@

```{r varviolin, fig.width=17, fig.height=7}
uncertainty::varviolin(in_var = in_var, out_var = out_var)
```

@@

`varkernelslicerange()` shows the influencing variable `in_var` value intervals, optimized interquartile ranges shown in `varviolin()` can be used to select a range to slice from the density kernel `varkernel()` as was done for a single variable value in `varkernelslice()`. Here we set the maximum expected value for `in_var` to `r (min(in_var)+(max(in_var)-min(in_var))/4*3)` and the minimum expected value to `r (min(in_var)+(max(in_var)-min(in_var))/4)`.
